---
title: "NMR metabolites and hypertension: XGBoost"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  html_document:
    number_sections: true
---

```{r options, include=FALSE}
knitr::opts_chunk$set(include = TRUE,
                      echo = TRUE,
                      message = FALSE,
                      results = 'asis',
                      cache = FALSE,
                      warning=FALSE,
                      error = FALSE)
options(error = traceback, knitr.kable.NA = "")
```

[Guide](https://michael-barr.com/datasci/bayesian-mbo)
[Fatty liver manuscript](https://academic.oup.com/eurheartj/article/41/44/4259/5899897#supplementary-data)
[Fatty liver source code](https://www.informahealthcare.com/doi/suppl/10.1080/19490976.2021.1888673/suppl_file/kgmi_a_1888673_sm1172.zip)

# Libraries

<details><summary>Libaries</summary>

```{r libraries, results = 'hide'}
library(magrittr)
library(knitr)
library(tidyr)
library(broom)
library(purrr)
library(dplyr)
library(xgboost)
library(caret)
library(mlr)
library(mlrMBO)
library(ggplot2)
library(vip)
library(parallel)
library(ggthemes)
library(ggrepel)
library(scales)
library(cowplot)
library(gridExtra)
library(optparse)
```

```{r set theme, include = FALSE}
theme_set(theme_tufte(base_family = "sans", base_size = 18) +
               theme(panel.border = element_rect(colour = "black", fill = NA),
                     axis.text = element_text(colour = "black", size = 18),
                     aspect.ratio = 1))
```

```{r commandline arguments, echo = FALSE}
options <- list(
    optparse::make_option(c("-n", "--no-clinical"),
                          action="store_true",
                          default = FALSE,
                          help="Exclude clinical covariates"),
    optparse::make_option(c("-m", "--no-metabolites"),
                          action="store_true",
                          default = FALSE,
                          help="Exclude metabolites covariates")
)

arguments <- parse_args(OptionParser(option_list = options))

exclude_vars <-  arguments[["no-clinical"]]
exclude_metabolites <-  arguments[["no-metabolites"]]

message("Args ", exclude_vars, " and ", exclude_metabolites)
```


</details>

## Functions

<details><summary>Helper functions</summary>

```{r, code = readLines("articles-xgfunctions.R")}
source("articles-xgfunctions.R")
```

</details>

<details><summary>XGBoost functions</summary>

```{r, code = readLines("articles-xgboost.R")}
source("articles-xgboost.R")
```

</details>

<details><summary>Function factory</summary>

```{r saver}
datetime <- format(Sys.time(), "%Y%m%d_%H%M%S")
mysave <- mysavefactory(arguments)
myload <- myloadfactory(arguments)
myggsave <- myggsavefactory(datetime)
datetime
```

</details>

#  Folder structure

```{r clean folders}
mkdir("cache", "report", "session")
```

# Data import

## Main data import

```{r import data}
dset <- readRDS("rds/dset-20210217-102531.rds")
```

## Covariates

Clinical covariates

```{r clinical variables, eval = !exclude_vars, echo = !exclude_vars}
ivars <- c("age", "female", "bmi", "smoker", "diabetes", "leisure", "bptreat", "koltreat")
ivars_longitudinal <- c("age", "female", "bmi", "smoker", "diabetes", "leisure", "bptreat", "koltreat", "sys_old")
```

```{r no clinical variables, eval = exclude_vars, echo = exclude_vars}
ivars <- c()
ivars_longitudinal <- c()
```

```{r metabolites, eval = !exclude_metabolites, echo = !exclude_metabolites}
metabolites.all <- getmetabolites(dset$crosssectional)
metabolites.subclass <- getsubclasses(metabolites.all)
metabolites <- metabolites.all %difference% metabolites.subclass
```


```{r no metabolic measures, eval = exclude_metabolites, echo = exclude_metabolites}
metabolites <- c()
```

## Data filtering


```{r modify dset for xgboost}
dset_all <- dset$crosssectional %>%
    mutate(fold = case_when(
               cohort == "1997" ~ 1,
               cohort == "T2000" ~ 2,
               cohort == "2002" ~ 3,
               cohort == "2007" ~ 4,
               cohort == "2012" ~ 5,
               cohort == "2017" ~ 6)) %>%
    mutate_at(ivars, as.numeric) %>%
    select(one_of(ivars), sys, one_of(metabolites), fold)

dset_longitudinal_all <- dset[["longitudinal"]] %>%
    get_longitudinal_dset %>%
    mutate_at(ivars_longitudinal, as.numeric) %>%
    select(one_of(ivars_longitudinal), cohort, sys, one_of(metabolites))
```



## Discovery and test sets

```{r splitting data}
dset_discovery <- filter(dset_all, fold %in% c(1, 2, 3, 4, 5))
dset_test <- filter(dset_all, fold == 6) 
```

Discovery dataset dimensions are `r dim(dset_discovery)` and test dataset dimensions are `r dim(dset_test)`.

```{r dset longitudinal}
dset_longitudinal_discovery <- dset_longitudinal_all %>% filter(cohort == "T2000") %>% select(-cohort) 
dset_longitudinal_test <- dset_longitudinal_all %>% filter(cohort == "F2007") %>% select(-cohort)
```

Longitudinal discovery dataset dimension are `r dim(dset_longitudinal_discovery)` and test dataset dimensions are
`r dim (dset_longitudinal_test)`

# XGBoost parameter optimization

```{r Bayesian optimization restore}
bayes_run <- myload("bayes_run")
longitudinal_bayes_run <- myload("longitudinal_bayes_run")
```


Defining basic xgboost-model parameters.

```{r basic parameters}
basic_xgboost_params <-  list( booster = "gbtree",
                              objective ="reg:squarederror",
                              eval_metric = "rmse")
```

Define model parameters ranges for cross-validation

```{r parameter set}
my_single_objective_params <- makeParamSet(
    makeIntegerParam("nrounds", lower = 50, upper = 5000),
    makeNumericParam("eta", lower = 0.001, upper = 0.1),
    makeNumericParam("gamma", lower = 0.1, upper = 100),
    makeIntegerParam("max_depth", lower= 2, upper = 10),
    makeIntegerParam("min_child_weight", lower= 1, upper = 200),
    makeNumericParam("subsample", lower = 0.2, upper = 0.8),
    makeNumericParam("colsample_bytree", lower = 0.2, upper = 0.8),
    makeIntegerParam("fold", lower = 1, upper = 5, tunable = FALSE))
```

```{r longitudinal parameter set}
my_longitudinal_single_objective_params <- makeParamSet(
    makeIntegerParam("nrounds", lower = 50, upper = 5000),
    makeNumericParam("eta", lower = 0.001, upper = 0.1),
    makeNumericParam("gamma", lower = 0.1, upper = 100),
    makeIntegerParam("max_depth", lower= 2, upper = 5),
    makeIntegerParam("min_child_weight", lower= 1, upper = 200),
    makeNumericParam("subsample", lower = 0.2, upper = 0.8),
    makeNumericParam("colsample_bytree", lower = 0.2, upper = 0.8))
```


Find optimal set of model parameters
	
```{r optimize parameters}
smoof_func <- smoof::makeSingleObjectiveFunction(
                         name = "xgb_cv_bayes",
                         fn =   function(x) my_leaveoneout_crossvalidation(x,
                                                                                               params = basic_xgboost_params,
                                                                                               dset = dset_discovery),
                         par.set = my_single_objective_params,
                         minimize = TRUE)
```

```{r longitudinal optimize parameters}
longitudinal_smoof_func <- smoof::makeSingleObjectiveFunction(
                                      name = "xgb_cv_bayes_longitudinal",
                                      fn =   function(x) my_n_fold_crossvalidation(x,
                                                                                                     params = basic_xgboost_params,
                                                                                                     dset = dset_longitudinal_discovery,
                                                                                                     nfolds = 5),
                                      par.set = my_longitudinal_single_objective_params,
                                      minimize = TRUE)
```


Run cross-validation between distinct cohorts: FR1997, FR2002 FR2007, FR2012,  FH2017, and H2000.

```{r extra params}
n_design <-  6*(get_nparameters(my_single_objective_params) - 1)
n_opt_steps <- 100
```

<details><summary>Optimization run</summary>

```{r Bayesian optimization, eval = is_null(bayes_run)}
bayes_run <- do_bayes(n_design = n_design,
                      of = smoof_func,
                      opt_steps = n_opt_steps)
mysave(bayes_run, name = "bayes_run")
```

```{r longitudinal Bayesian optimization, eval = is_null(longitudinal_bayes_run)}
longitudinal_bayes_run <- do_bayes(n_design = n_design,
                                   of = longitudinal_smoof_func,
                                   opt_steps = n_opt_steps)
mysave(longitudinal_bayes_run, name = "longitudinal_bayes_run")
```


</details>

<details><summary>Diagnostic plots</summary>

```{r optimization graph, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
bayes_run$opt.path$env$path %>%
    mutate(Round = row_number()) %>%
    mutate(type = ifelse(Round <= n_design, "Design", "mlrMBO optimization")) %>%
    ggplot(aes(x= Round, y= y, color= type)) + 
    geom_point(size = 1) +
    scale_x_continuous(expand = c(0, 0), name = "Round") +
    scale_y_continuous(expand = c(0, 0), name = "-log(likelihood)", limit = c(0, 100)) +
    labs(title = "mlrMBO optimization") +
    theme_classic()
```

```{r longitudinal optimization graph, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
longitudinal_bayes_run$opt.path$env$path %>%
    mutate(Round = row_number()) %>%
    mutate(type = ifelse(Round <= n_design, "Design", "mlrMBO optimization")) %>%
    ggplot(aes(x= Round, y= y, color= type)) + 
    geom_point(size = 1) +
    scale_x_continuous(expand = c(0, 0), name = "Round") +
    scale_y_continuous(expand = c(0, 0), name = "-log(likelihood)", limit = c(0, 100)) +
    labs(title = "mlrMBO optimization") +
    theme_classic()
```


```{r basic graphs, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
plot(bayes_run)
```

```{r basic longitudinal graphs, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
plot(longitudinal_bayes_run)
```


</details>

## Optimal xgboost models

Define optimal params for xgboost

```{r best cv model}
( my_optimal_params <- modifyList(basic_xgboost_params, bayes_run[["x"]])) %>%
    map_df(~.x) %>%
    kable
```

```{r best longitudinal cv model}
( my_optimal_longitudinal_params <- modifyList(basic_xgboost_params, longitudinal_bayes_run[["x"]])) %>%
    map_df(~.x) %>%
    kable
```

```{r run xgboost with optimal params}
set.seed(42)
my_optimal_model <- my_optimal_xboost(params = my_optimal_params %>%
                                          purrr::list_modify(nrounds = NULL, fold = NULL),
                                      dset = dset_discovery,
                                      nrounds = my_optimal_params[["nrounds"]])
```

```{r run longitudinal xgboost with optimal params}
set.seed(42)
my_optimal_longitudinal_model <- my_optimal_xboost(params = my_optimal_longitudinal_params %>%
                                                       purrr::list_modify(nrounds = NULL, fold = NULL),
                                      dset = dset_longitudinal_discovery,
                                      nrounds = my_optimal_longitudinal_params[["nrounds"]])
```

## Feature importance


<details><summary>Numerical cross-sectional values</summary>

```{r importance table}
xgb.importance(model = my_optimal_model) %>% kable
```

</details>

<details><summary>Numerical longitudinal values</summary>

```{r longitudinal importance table}
xgb.importance(model = my_optimal_longitudinal_model) %>% kable
```

</details>


## Prediction

### Cross-sectional

#### Training set

```{r training prediciton}
list(xsectional_training = summarize_run(dset_discovery, my_optimal_model),
     xsectional_test = summarize_run(dset_test, my_optimal_model),
     longitudinal_training = summarize_run(dset_longitudinal_discovery, my_optimal_longitudinal_model),
     longitudinal_test = summarize_run(dset_longitudinal_test, my_optimal_longitudinal_model)) %>%
    map_df(identity, .id = "model") %>%
    kable()
```

<details><summary>Prediction plots</summary>

```{r training prediction plot, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
summarize_run(dset_discovery, my_optimal_model, plot = TRUE)
```

```{r testing prediction plot, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
summarize_run(dset_test, my_optimal_model, plot = TRUE)
```

```{r longitudinal training prediction plot, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
summarize_run(dset_longitudinal_discovery, my_optimal_longitudinal_model, plot = TRUE)
```
```

```{r longitudinal testing prediction plot, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
summarize_run(dset_longitudinal_test, my_optimal_longitudinal_model, plot = TRUE)
```

</details>


## Define variable names

```{r vip colnames}
new_colnames <- human_readable_names(dset_discovery %>% select(-sys, -fold), length = 12) %>% tibble::deframe()
```

## Variable importance in projection

```{r max vip}
xgboost_max_gain <- get_xgb_top_gain(my_optimal_model, my_optimal_longitudinal_model) %>%
    round_up_decimal
```

### Cross-sectional

```{r vip plot}
g_vip <- vip(my_optimal_model, num_features = 15) + 
    scale_x_discrete(labels= new_colnames, expand = c(0, 0)) +
    scale_y_continuous(limit = c(0, xgboost_max_gain))  +
    theme_classic() +
    theme(text = element_text(size=20))

myggsave(file = "cache/vip.png",
       plot = g_vip,
       width = 6,
       height = 3,
       dpi = 300,
       unit = "in")
```

### Longitudinal

```{r vip colnames longitudinal}
new_longitudinal_colnames <- human_readable_names(dset_longitudinal_discovery %>% select(-sys)) %>% tibble::deframe()
```

```{r vip plot longitudinal}
g_vip_longitudinal <- vip(my_optimal_longitudinal_model, num_features = 15) + 
    scale_x_discrete(labels= new_longitudinal_colnames) +
    scale_y_continuous(limit = c(0, xgboost_max_gain))  +
    theme_classic() +
    theme(text = element_text(size=20))

myggsave(file = "cache/vip_longitudinal.png",
       plot = g_vip_longitudinal,
       width = 20,
       height = 10,
       dpi = 300,
       unit = "in") 
```

## Partial dependency

### Important measures

```{r get most important NMR features}
important_crosssectional_measures <- xgb.importance(model = my_optimal_model) %>%
    arrange(-Gain) %>%
    filter(grepl("^NMR_", Feature)) %>%
    head(n = 9) %>%
    pull(Feature)
```

```{r get most important longitudinal NMR features}
important_longitudinal_measures <- xgb.importance(model = my_optimal_longitudinal_model) %>%
    arrange(-Gain) %>%
    filter(grepl("^NMR_", Feature)) %>%
    head(n = 9) %>%
    pull(Feature)
```

### Cross-sectional

```{r cross-sectional pdp matrix}
matrix_pdp <- dset_discovery %>% select(-any_of(c("sys", "fold"))) %>% data.matrix
```

```{r partial dependency}
df_pdp <- c2l(important_crosssectional_measures) %>%
    map(~ pdp::partial(my_optimal_model,
                       pred.var = .x,
                       pred.grid = crossing(!!quo_name(.x) := seq(-2, 2, 0.1)),
                       train = matrix_pdp,
                       ice = FALSE,
                       type = "regression",
                       plot = FALSE) %>% 
            as_tibble %>%
          rename(y = yhat, x := !!quo_name(.x))) %>%
    map_df(~.x, .id = "feature")
```


### Longitudinal

```{r longitudinal pdp matrix}
matrix_longitudinal_pdp <- dset_longitudinal_discovery %>% select(-any_of(c("sys", "fold"))) %>% data.matrix
```

```{r partial dependency longitudinal}
df_pdp_longitudinal <- c2l(important_longitudinal_measures) %>%
    map(~ pdp::partial(my_optimal_longitudinal_model,
                       pred.var = .x,
                       pred.grid = crossing(!!quo_name(.x) := seq(-2, 2, 0.1)),
                       train = matrix_longitudinal_pdp,
                       type = "regression",
                       ice = FALSE,
                       plot = FALSE)  %>% 
            as_tibble %>%
          rename(y = yhat, x := !!quo_name(.x))) %>%
    map_df(~.x, .id = "feature")
```

### Prepare ggplot graphs

```{r plot cross-sectional pd, echo = FALSE, eval = !exclude_metabolites}
g_pdp <- df_pdp %>%
    mutate(desc = new_colnames[feature]) %>%
    ggplot(aes(x = x, y = y)) +
    facet_wrap(~desc, labeller = label_wrap_gen(width = 14)) +
    geom_line() +
    scale_x_continuous(breaks = seq(-2,2,1),
                       limits = c(-2, 2)) +
    scale_y_continuous(breaks = seq(110, 180, 2)) +
    xlab("1-SD change in metabolic measure") +
    ylab("Systolic BP") +
    theme(strip.text = element_text(size = 14),
          axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)))

myggsave(file = "cache/pdp.png",
           plot = g_pdp, 
           width = 9,
           height = 8,
           dpi = 300,
        unit = "in")
```


```{r plot longitudinal pd, echo = FALSE, eval = !exclude_metabolites}
g_pdp_longitudinal <-  df_pdp_longitudinal %>%
    mutate(desc = new_colnames[feature]) %>%
    ggplot(aes(x = x, y = y)) +
    facet_wrap(~desc, labeller = label_wrap_gen(width = 24)) +
    geom_line() +
    scale_x_continuous(breaks = seq(-2,2,1),
                       limits = c(-2, 2)) +
    scale_y_continuous(breaks = seq(110, 180, 1)) +
    xlab("1-SD change in metabolic measure") +
    ylab("Follow-up systolic BP")  +
    theme(strip.text = element_text(size = 14),
          axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)))

myggsave(file = "cache/pdp_longitudinal.png",
           plot = g_pdp_longitudinal , 
           width = 10,
           height = 8,
           dpi = 300,
        unit = "in")
```

## Generate images

```{r join pdp images, eval = !exclude_metabolites}
list(g_pdp, g_pdp_longitudinal) %>%
    map(fix_ggplot_labels) %>%
    plot_grid(plotlist = .,
              labels = "AUTO",
              label_fontfamily = "sans",
              label_fontface = "plain",
              label_size = 30) %>%
    myggsave(file = "cache/pdp_all.png",
             plot = .,
             width = 14,
             height = 8,
             dpi = 300,
             unit = "in")
```

```{r join vip images}
plot_grid(plotlist = normalize_plots(g_vip, g_vip_longitudinal),
          ncol = 2,
          labels = "AUTO",
          label_fontfamily = "sans",
          label_fontface = "plain",
          label_size = 30) %>%
    myggsave(file = "cache/vip_all.png",
           plot = .,
           width = 14,
           height = 6,
           dpi = 300,
           unit = "in")
```

