---
title: "NMR metabolites and hypertension: Data handling"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
---

```{r options, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = FALSE, results='asis', warning=FALSE)
now <- format(Sys.time(), '%Y%m%d-%H%M%S')
```

# Libraries

<details>
  <summary>Libraries</summary>

```{r libraries, results = 'hide'}
library(magrittr)
library(tibble)
library(readr)
library(knitr)
library(kableExtra)
library(tidyr)
library(broom)
library(dplyr)
library(parallel)
library(officer)
library(flextable)
library(ggplot2)
library(randomForest)
library(Boruta)
library(pander)
library(caret)
library(pROC)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(haven)
library(readxl)
library(openxlsx)
library(ggpubr)
library(purrr)
library(ggforestplot)
library(xfun)
```

</details>

## Functions

<details><summary>Functions</summary>

```{r Functions}
sourcefiles <- c("articles-functions.R",
                 "articles-plots.R",
                 "articles-importer.R",
                 "articles-definitions.R",
                 "articles-tables.R")
```

```{r import files, echo = FALSE}
for (f in sourcefiles) {
    source(f)
}
```

```{r embbed files, echo = FALSE}
xfun::embed_files(c("rrnmr-import.Rmd", sourcefiles))
```

</details>


# Variable definitions

Included variables

```{r names}
vars <- c("age", "bmi", "bptreat", "cohort", "diabetes", "dias",
          "female", "htn", "htn3", "koltreat", "leisure", "sampleid",
          "smoker", "sys")
```

Excluded metabolites

```{r Metabolites}
exclude.metabolites <- excludemetabolites(byname = c("NMR_HDL2_C",
                                                     "NMR_HDL3_C",
                                                     "NMR_ApoB_by_ApoA1",
                                                     "NMR_TG_by_PG",
                                                     "NMR_Free_C",
                                                     "NMR_Esterified_C",
                                                     "NMR_Remnant_C",
                                                     "NMR_DHA",
                                                     "NMR_LA",
                                                     "NMR_MUFA",
                                                     "NMR_Omega_3",
                                                     "NMR_Omega_6",
                                                     "NMR_PUFA",
                                                     "NMR_SFA"),
                                          bygroup = c("Relative lipoprotein lipid concentrations",
                                                      "Lipoprotein subclasses"))
```

<details>
  <summary>List of excluded metabolites</summary>

```{r excluded metabolites list, echo = FALSE}
data.frame(metabolite = exclude.metabolites) %>%
    mutate(nmrgroup = bioproperty(metabolite, "group")) %>%
               mykable
```

</details>

# Cross-sectional sample

## Data import

Reading phenotype and NMR data for *HEALTH 2000*.

```{r Reading T2000}
dset.t2000 <- HEALTH_definitions("nmrdata/T20xx/T2000-phenotype.sas7bdat",
                                 "nmrdata/T20xx/T2000-phenotype-extra.sas7bdat",
                                 "nmrdata/T20xx/T2000-nmr.sas7bdat",
                                 vars,
                                 cohort = "T2000") %>%
    select(-SAMPLEID, -one_of(exclude.metabolites))
```

Reading phenotype data for *FINRISK 1997-2012*.

```{r read sas data for FR/FT}
dset.finrisk.pheno <- my_read_sas("nmrdata/fr2019_015_ft2019_016_nmr_10_23.sas7bdat") %>%
    dplyr::select(-one_of(metabolitemapping(prefix = ""))) %>%
    dplyr::filter(VUOSI != 1992)
```

Reading NMR-data for *FINRISK*.

```{r Reading NMR-data for FR}
dset.finrisk.nmr <- c("nmrdata/FR-NMR/nmr_fr97.sas7bdat",
                      "nmrdata/FR-NMR/nmr_fr02.sas7bdat",
                      "nmrdata/FR-NMR/nmr_fr07.sas7bdat",
                      "nmrdata/FR-NMR/nmr_fr12.sas7bdat",
                      "nmrdata/FR-NMR/nmr_ft17.sas7bdat") %>%
    purrr::map(my_read_nmr) %>%
    map_df(., ~.x) %>%
    mutate(sampleid = sprintf("%s_%s", VUOSI, HAVTUN))  %>%
    select(-one_of(exclude.metabolites))
```


Joining *FINRISK* phenotype and NMR data, and biding *FINRISK* and *HEALTH 2000*.
When participant has taken part in more than one *FINRISK* examination,
we exclude all but the latest one.

```{r crossectional missing values}
dset.crosssectional.raw <- dset.finrisk.pheno %>%
    remove_reexamined %>%
    FR_definitions(variables = vars) %>%
    full_join(., dset.finrisk.nmr, by = "sampleid") %>%
    bind_rows(., dset.t2000) %>%
    filter(!is.na(cohort))
```

<details>
	<summary>Missing values</summary>

Missing values

```{r cross sectional missing by cohort, results='asis', echo = FALSE}
missingbycohort(dset.crosssectional.raw, vars) %>% mykable
```

Metabolite values bofre normalization

```{r cross sectional nmr table, echo = FALSE}
dset.crosssectional.raw %>% select(starts_with("NMR_")) %>% mydescribe %>% mykable
```

</details>

## Finalizing data frame

Excluding participants with missing values.

```{r crossectional finalizing}
dset.crosssectional <- dset.crosssectional.raw %>%
    select(-HAVTUN, -VUOSI) %>%
    filter_at(.vars = vars(vars), all_vars(!is.na(.))) %>%
    scale_and_filter_nmr %>%
    mutate(cohort = as.factor(cohort),
           agegroup = factor(ifelse(age < median(age), "young", "old")))
```

```{r crossectional characteristics, echo = FALSE}
characteristicsTable(dset.crosssectional, strata = "cohort") %>% mykable
```

<details>
	<summary>Missing values</summary>

Missing values 

```{r cross sectional missing by cohort after finalizing, results='asis', echo = FALSE}
missingbycohort(dset.crosssectional, vars) %>% mykable
```
	
Raw number of participants by NMR data

```{r Number of participans by available NMR-data, echo = FALSE}
dset.crosssectional %>%
    select(cohort, starts_with("NMR_")) %>%
    mutate_at(vars(starts_with("NMR_")), is.na) %>%
    group_by(cohort)  %>%
    summarize_all(sum) %>%
    gather(metabolite, value, -cohort) %>% 
    spread(cohort, value) %>%
    mykable
```

</details>


## Diagnostic plots

Box plot for metadata

```{r scatter plots for metadata, fig.width=8, fig.height=4, echo = FALSE}
myboxplot(dset.crosssectional, vars = c("age", "bmi", "sys", "dias"), normalize = TRUE)
```

Box plot for metabolites

```{r scatter plots for metabolites, fig.width=20, fig.height=4, echo = FALSE}
myboxplot(dset.crosssectional,
          vars = dset.crosssectional %>% colnames %>% mygrep(word = "NMR_"),
          normalize = TRUE,
          rename = TRUE)
```

```{r qq plot for crosssectional data,  fig.width=20, fig.height=10}
diagnosticqqplot(dset.crosssectional, dset.crosssectional %>% colnames %>% mygrep(word = "NMR_"))
```

# Longitudal sample

## DILGOM

*DILGOM* phenotype data is included in the columns of *FINRISK 2007* and can be parsed.

```{r dilgom}
dilgom.baseline <- FR_definitions(dset.finrisk.pheno %>% filter(VUOSI == 2007), variables = vars)
dilgom.followup <- DILGOM_definitions(dset.finrisk.pheno %>% filter(VUOSI == 2007), variables = vars)
```

N:o FR07 participants with miggins DILGOM systole measurement
`r dset.finrisk.pheno %>% filter(VUOSI == 2007) %>% pull(DLGM14_VP_OMRON1_SYS) %>% is.na %>% sum`.


```{r Reading NMR-data for DILGOM}
dset.dilgom.nmr <- c("nmrdata/FR-NMR/nmr_fr07.sas7bdat",
                     "nmrdata/FR-NMR/nmr_dilgom14.sas7bdat") %>%
    purrr::map(my_read_nmr) %>%
    purrr::reduce(rbind) %>%
    mutate(sampleid = sprintf("%s_%s", VUOSI, HAVTUN)) %>%
    select(-one_of(exclude.metabolites))
```

Joining phenotype and NMR-data

```{r dilgom join pheno and nmr data}
dset.dilgom <- rbind(dilgom.baseline %>% mutate(cohort = "F2007"),
                     dilgom.followup %>% mutate(cohort = "F2014")) %>%
    right_join(., dset.dilgom.nmr, by = "sampleid") %>%
    select(-HAVTUN, -VUOSI) %>%
    mutate(sampleid = gsub("^2007_", "2014_", sampleid)) %>%
    filter(!is.na(cohort))
```

## HEALTH

Reading phenotype and NMR data for *HEALTH 2011* cohort. Number of
follow-up pairs found between samples.  SAS object for NMR data in
T2000 has `r read_sas("nmrdata/T20xx/T2000-nmr.sas7bdat") %>% nrow`
and in T2011 `r read_sas("nmrdata/T20xx/T2011-nmr.sas7bdat") %>% nrow`
rows.

```{r read sas data for T2011}
dset.t2011 <- HEALTH_definitions("nmrdata/T20xx/T2011-phenotype.sas7bdat",
                                 "nmrdata/T20xx/T2011-phenotype-extra.sas7bdat",
                                 "nmrdata/T20xx/T2011-nmr.sas7bdat",
                                 vars,
                                 cohort = "T2011") %>%
    select(-one_of(exclude.metabolites))
```

## Combining *DILGOM* and *HEALTH*

Binding DILGOM and HEALTH studies.

```{r dilgom cleaned}
dset.longitudinal.raw <- rbind(dset.dilgom,
                               dset.t2000,
                               dset.t2011)
```

<details>
	<summary>Missing values</summary>

Missing values 

```{r dilgom missing by cohort, echo = FALSE}
missingbycohort(dset.longitudinal.raw, vars) %>% mykable
```

</details>

## Finalizing data frame

Excluding participants with missing values and those with missing baseline-followup data.

```{r longitudinal finalizing}
dset.longitudinal <- dset.longitudinal.raw %>%
    filter_at(vars(starts_with("NMR")), any_vars(!is.na(.))) %>%
    mutate_at(vars(starts_with("NMR")), ~scale(.)) %>%
    mutate(agegroup = factor(ifelse(age < median(age), "young", "old"))) %>%
    cleanlongitudinal()
```

The number of participants before and after clean

```{r longitudinal characteristics, echo = FALSE}
characteristicsTable(dset.longitudinal, strata = "cohort") %>% mykable
```

## Diagnostic graphs

Box plot for metadata

```{r scatter plots for metadata longitudinal, fig.width=8, fig.height=4, echo = FALSE}
myboxplot(dset.longitudinal, vars = c("age", "bmi", "sys", "dias"), normalize = TRUE)
```

Box plot for metabolites

```{r scatter plots for metabolites longitudinal, fig.width=20, fig.height=4, echo = FALSE}
myboxplot(dset.longitudinal, 
          vars = dset.longitudinal %>% colnames %>% mygrep(word = "NMR_"),
          normalize = TRUE,
          rename = TRUE)
```


# Saving data frame for analyzes

```{r save data}
dir.create("rds", showWarnings = FALSE)
saveRDS(list("crosssectional" = dset.crosssectional,
             "longitudinal" = dset.longitudinal),
        file = sprintf("rds/dset-%s.rds", now))
```

